---
title: "DE_pseudobulk"
author: "Heesoo Song"
date: '2021 12 1 '
output: html_document
---

Extension of *DE_EdgeR.Rmd*. Here, we will create aggregate matrix to observe change spanning over celltype rather than time point.
https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf

```{r}
#install.packages('Matrix.utils')
#install.packages("statmod")
#BiocManager::install("org.Mm.eg.db")
```


```{r, message = FALSE}
library(SeuratObject)
library(Seurat)
library(nichenetr)
library(tidyverse)
library("ggplot2")
library(muscat)
library(SingleCellExperiment)
library(purrr)
library(Matrix.utils)
library(RColorBrewer)
library(pheatmap)
library(cluster)
library(celda)
```

```{r}
# Load processed expression data of interacting cells
hepatectomy_expression = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/seurat_obj_phx_25102021.rds")

# Total 91021 cells, with 5000 genes
```

```{r}
celltype_OI <- c("cholangiocyte","endothelial","hepatocyte","kupffer","monocyte","neutrophil","stellate")
# experiments <- c("0_phase_1", "3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_1", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")
experiments <- c("3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")

#hours <- c(0, 3, 6, 12, 24, 36, 48, 48, 72, 96, 120, 168)
hours <- c(3, 6, 12, 24, 36, 48, 72, 96, 120, 168)
```

```{r}
hepa_celltypeOI = subset(x = hepatectomy_expression, idents = celltype_OI)

hepa_celltypeOI@meta.data$time_experiment <- factor(hepa_celltypeOI@meta.data$time_experiment, levels = experiments)

hepa_celltypeOI = subset(x = hepa_celltypeOI, subset = time_experiment %in% experiments)
```

```{r}
rm(hepatectomy_expression)
```

********************************************************************************
**WARNING: Skip this chunk if you already have decontX sce object**

# Create Aggregate Count Matrix

```{r}
# sce <- as.SingleCellExperiment(hepa_celltypeOI)
```

```{r}
# ## Determine the number of cells per sample
# table(sce$time_experiment)
```

```{r}
# counts(sce)[1:6,1:4]
```

```{r}
# # library(biomaRt)
# # ensembl <- useEnsembl(biomart="ensembl", dataset = "mmusculus_gene_ensembl")
# # IDtoSymbol_matrix <- getBM(attributes=c("ensembl_gene_id","external_gene_name"), filters = "ensembl_gene_id", values = unique(rownames(sce)), mart = ensembl)
# 
# IDtoSymbol_matrix <- read.csv("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/IDtoSymbol_matrix.csv")
```

```{r}
# rownames_sce <- data.frame("ENSEMBL" = rownames(sce))
# rownames_sce["order"] <- c(1:dim(rownames_sce)[1])
# gene_ID <- merge(rownames_sce, IDtoSymbol_matrix, by.x="ENSEMBL", by.y = "ensembl_gene_id", all.x=TRUE)
# 
# # Fill rest of NA with ENSEMBL ID (ID no longer in the ENSEMBL database)
# gene_ID$external_gene_name[which(gene_ID$external_gene_name == "")] <- gene_ID$ENSEMBL[which(gene_ID$external_gene_name == "")]
# gene_ID$external_gene_name[which(is.na(gene_ID$external_gene_name))] <- gene_ID$ENSEMBL[which(is.na(gene_ID$external_gene_name))]
# gene_ID <- gene_ID[order(gene_ID$order),]
# 
# # Change rownames of SCE
# rownames(sce) <- gene_ID$external_gene_name
# 
# counts(sce)[1:6,1:4]

```

## Remove Ambient RNA
```{r message = FALSE}
# library(celda)
```

```{r}
# # calculate contamination
# # WARNING: Long runtime - took 1.318807 hours
# sce <- decontX(sce)
```

```{r}
# # DecontX UMAP designed to find "broad" cell types
# umap <- reducedDim(sce, "decontX_UMAP")
# plotDimReduceCluster(x = sce$decontX_clusters,
#                      dim1 = umap[, 1], dim2 = umap[, 2])
```

```{r}
# # Contamination on UMAP
# # % of contamination in each cell. Visualize clusters with higher level of ambient RNA
# 
# plotDecontXContamination(sce)
```

```{r}
# library(scater)
# sce <- logNormCounts(sce)
```


```{r}
# # Expression of markers on UMAP
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Csf1r", "Sparc"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "problematic genes")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Gjb2", "C8g"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Hepatocyte Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Spp1", "Epcam"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Cholangiocyte Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Ccr2"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Monocyte Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Cd5l", "Marco"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "KC_Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("S100a8", "S100a9"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Neutrophil Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Dcn", "Col3a1"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Stellate Markers")
# 
# plotDimReduceFeature(as.matrix(logcounts(sce)),
#     dim1 = umap[, 1],
#     dim2 = umap[, 2],
#     features = c("Egfl7", "Tie1"),
#     ncol = 2,
#     exactMatch = TRUE) + labs(title = "Endothelial Markers")
```

```{r}
# Plot_features <- function(gene, seurat_obj, geneID_convert) {
#   
#   # Change gene symbol to ENSEMBL to subset seurat object
#   ENSEMBL <- geneID_convert$ensembl_gene_id[which(geneID_convert$external_gene_name == gene)]
# 
#   # UMAP highlighting cells with geneOI
#   feature_baseplot <- FeaturePlot(seurat_obj, features = ENSEMBL, min.cutoff = "q10", max.cutoff = "q90" ,label = TRUE) + labs(title = gene)
#   
#   print(feature_baseplot)
# }
```

```{r, message=FALSE}
# #setwd("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet")
# 
# for (gene in c("Gjb2", "C8g", "Spp1", "Epcam", "Ccr2", "Cd5l", "Marco", "S100a8", "S100a9", "Dcn", "Col3a1", "Egfl7", "Tie1")){
#   Plot_features(gene, hepa_celltypeOI, IDtoSymbol_matrix)
# }
```

```{r}
# # Plotting error occurs when assigning various cluters for one cell type. This is solved in Github version but not in Bioconductor (which I had installed with)
# # So based on the cluster we have through DecontX, we have to regenerate SCE with new cluster number.
# # WHICH AGAIN TAKES AN HOUR LONG PROCESS WTF
# newz <- colData(sce)$decontX_clusters
# newz[newz == 4] = 3 #Endothelial
# newz[newz %in% c(5,6,7,8)] = 4 # Hepatocyte
# newz[newz %in% c(9,10,12)] = 5 # Kupffer
# newz[newz == 11] = 6 # Monocyte
# newz[newz %in% c(13,14)] = 7 # Neutrophil
# 
# sce <- decontX(sce, z = newz)
```

********************************************************************************

```{r}
#saveRDS(sce, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/sce_DecontX.rds")
sce <- readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/sce_DecontX.rds")
```

```{r}
# DecontX UMAP designed to find "broad" cell types
umap <- reducedDim(sce, "decontX_UMAP")
plotDimReduceCluster(x = sce$decontX_clusters,
                     dim1 = umap[, 1], dim2 = umap[, 2])
```

```{r}
# Barplot of markers detected in cell clusters

markers <- list(Hepatocyte_Markers = c("Gjb2", "C8g"),
    Cholangiocyte_Markers = c("Spp1", "Epcam"),
    Monocyte_Markers = c("Ccr2"),
    KC_Markers = c("Cd5l", "Marco"),
    Neutrophil_markers = c("S100a8", "S100a9"),
    Endothelial_markers = c("Egfl7", "Tie1"),
    Stellate_markers = c("Dcn", "Col3a1"))


# This form of groupCluster works in new Github version. But not Bioconductor version.
# cellTypeMappings <- list(Hepatocyte = c(5, 6, 7, 8), Cholangiocyte = 1, Monocyte = 11, Kupffer = c(9, 10, 12),
#                          Neutrophil = c(13, 14), Stellate = 2, Endothelial = c(3, 4))


cellTypeMappings <- list(Hepatocyte = 4, Cholangiocyte = 1, Monocyte = 6, Kupffer = 5,
                         Neutrophil = 7, Stellate = 2, Endothelial = 3)

plotDecontXMarkerPercentage(sce,
    markers = markers,
    groupClusters = cellTypeMappings,
    assayName = c("counts", "decontXcounts"),
    labelSize = 2)
```

```{r}
# Violin Plot
Markers <- c("Hepatocyte_Markers","Cholangiocyte_Markers","Monocyte_Markers",
      "KC_Markers","Neutrophil_markers","Endothelial_markers","Stellate_markers")
for (marker in Markers){
  plotDecontXMarkerExpr <- plotDecontXMarkerExpression(sce,
      markers = markers[[marker]],
      groupClusters = cellTypeMappings,
      ncol = 4) + labs(title = marker)
  print(plotDecontXMarkerExpr)
}
```


```{r}
# Contamination level on Original UMAP
umap_df <- data.frame(hepa_celltypeOI[["umap"]]@cell.embeddings, "Contamination" = colData(sce)$decontX_contamination, "Celltype" = sce$ident)
umap_df <- umap_df[order(umap_df$Contamination), ]
head(umap_df)
```
```{r}
label <- aggregate(umap_df[, 1:2], list(umap_df$Celltype), mean)
label
```

```{r}
Decont_orig_umap <- ggplot2::ggplot(
    umap_df,
    ggplot2::aes_string(
      x = colnames(umap_df)[1],
      y = colnames(umap_df)[2]
    )
  ) +
    ggplot2::geom_point(
      stat = "identity",
      size = 1,
      ggplot2::aes_string(color = "Contamination")
    ) +
    ggplot2::theme_bw() +
    ggplot2::scale_colour_gradientn(
      colors = c("blue","green","yellow","orange","red"),
      name = "Contamination",
      limits = c(0, 1)
    ) +
    ggplot2::theme(
      panel.grid.major = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      axis.text = ggplot2::element_text(size = 15),
      axis.title = ggplot2::element_text(size = 15)
    ) +
    ggplot2::geom_text(data = label, aes(label = Group.1))
print(Decont_orig_umap)
```


## Create aggregate matrix
```{r}
# Identify groups for aggregation of counts
#groups <- colData(sce)[, c("cluster_id", "time_experiment")]
groups <- colData(sce)[, c("ident", "time_experiment")]
```

```{r}
# Create aggregate matrix
# pb <- aggregate.Matrix(t(counts(sce)),
#                        groupings = groups, fun = "sum")
pb <- aggregate.Matrix(t(decontXcounts(sce)),
                       groupings = groups, fun = "sum")

dim(pb)
pb[1:6, 1:6]
```

```{r}
# Not every cluster is present in all samples; create a vector that represents how to split samples
# But actually we do have all cell types! :) Maybe skip this step
splitf <- sapply(stringr::str_split(rownames(pb), pattern="_", n=2), '[', 1)
```

```{r}
# Turn into a list and split the list into components for each cluster and transform, so rows are genes and columns are samples and make rownames as the sample IDs
library(magrittr)
pb <- split.data.frame(pb, 
                       factor(splitf)) %>%
  lapply(function(u) 
    set_colnames(t(u), 
                 stringr::str_extract(rownames(u), "(?<=_)[:alnum:]+_phase_[0-9]"))) 


# Explore the different components of list
names(pb)
```

```{r}
pb[[1]][1:6, 1:6]
```

```{r}
#Change row names into geneID + celltype
for (i in 1:length(names(pb))){
  rownames(pb[[i]]) <- paste0(rownames(pb[[i]]), "_", names(pb)[i])
}

pb[[1]][1:6, 1:6]
```


```{r}
# Combine matrices
pb_combined <- pb[[1]]
dim(pb_combined)

for (i in 2:length(names(pb))) {
  pb_combined <- rbind(pb_combined, pb[[i]])
}
dim(pb_combined)
cluster_counts <- as.data.frame(as.matrix(pb_combined))
```

```{r}
# extract_geneID <- as.data.frame(str_split(rownames(cluster_counts), "[.]", simplify=TRUE))
# cluster_counts$ENSEMBL <- extract_geneID$V1
# cluster_counts$celltype <- extract_geneID$V2
# 
# library(biomaRt)
# ensembl <- useEnsembl(biomart="ensembl", dataset = "mmusculus_gene_ensembl")
# IDtoSymbol_matrix <- getBM(attributes=c("ensembl_gene_id","external_gene_name"), filters = "ensembl_gene_id", values = unique(extract_geneID$V1), mart = ensembl)
# 
# cluster_counts <- merge(cluster_counts, IDtoSymbol_matrix, by.x="ENSEMBL", by.y = "ensembl_gene_id", all.x=TRUE)
# colnames(cluster_counts)[length(colnames(cluster_counts))] <- "Symbol"
# 
# # Fill rest of NA with ENSEMBL ID (ID no longer in the ENSEMBL database)
# cluster_counts$Symbol[which(is.na(cluster_counts$Symbol))] <- cluster_counts$ENSEMBL[which(is.na(cluster_counts$Symbol))]
# cluster_counts$Symbol[which(cluster_counts$Symbol == "")] <- cluster_counts$ENSEMBL[which(cluster_counts$Symbol == "")]
# 
# rownames(cluster_counts) <- paste0(cluster_counts$Symbol, "_", cluster_counts$celltype)
# cluster_counts <- subset(cluster_counts, select=-c(ENSEMBL, celltype, Symbol)) ####
# 
# head(cluster_counts)
```


```{r}
# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$ident, sce$time_experiment)
```

# 2. Differential gene expression
The aim is to examine the relationship between gene abundances and time points per each cell-type. In other words, we are looking for genes that change expression level over time in a time course experiment. Given enough number of time points, it is reasonable to assume smooth changes in gene expression. In this way, polynomial or a cubic spline curve can be used to model gene expression along time with a certain number of degrees of freedom.


Perform DE analysis

Explore aggregated count data by MDS (multi-dimensional scaling) plots. This visualizes the differences between the expression profiles of different samples (cholangiocyte) in two dimensions.
```{r}
library(edgeR)
```
```{r}
plotMDS(cluster_counts, labels=colnames(cluster_counts))
```

**PCA per cell type**
```{r}
for (i in 1:length(pb)){
  colnames(pb[[i]]) <- hours
}

for (i in 1:length(pb)){
  celltypeMDS <- plotMDS(pb[[i]], labels=colnames(pb[[i]]), main = names(pb[i]))
  print(celltypeMDS)
}
```

Use cubic regression spline curve with 4 degrees of freedom. The advantage of using a cubic spline curve is that it provides more stable fit at the end points compared to a polynomial. The spline curve with 3 degrees of freedom has 2 knots where cucbic polynomicals are splined together.
```{r}
library(splines)
X <- ns(hours, df = 4)
design <- model.matrix(~X)
design
```
Estimate dispersion. The negative binomial (NB) dispersion is estimated. This returns the **DGEList** object with additional entries for the estimated NB dispersion for each gene.
```{r}
y <- DGEList(counts = cluster_counts, group = hours)
```

```{r}
# filter out lowly expressed genes
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes = FALSE]

# TMM Normalisation
# y <- calcNormFactors(y)
# y$samples
```


```{r}
y <- estimateDisp(y, design)
sqrt(y$common.dispersion)

# plot the tagwise dispersion against log2-CPM (counts per million)
plotBCV(y)
```


Fit a quasi-likelihood NB generalized log-linear model

```{r}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```

Fit a quasi-likelihood negative binomial generalized log-linear model to count data.
```{r}
fit <- glmQLFTest(fit, coef=2:5)
```

Multiple testing across genes and contrasts. The total number of genes with significant (5% FDR) changes at different time points. 
All spline coefficients are tested together. Thus t-tests for the individual coefficients and the logFC values do not have any interpretable meaning.

```{r}
sig_num <- summary(decideTests(fit))
sig_num
```

```{r}
# Genes with most significant time effects
# H0: no trend/pattern in expression level over time
tab <- as.data.frame(topTags(fit, n=dim(fit)[1]))
head(tab)
```

LogFC = log fold change, log difference between groups.
LogCPM = log count per million, measure of expression level.


```{r}
logCPM.obs <- edgeR::cpm(y, log=TRUE, prior.count=fit$prior.count)
logCPM.fit <- edgeR::cpm(fit, log=TRUE)
```

```{r}
# write.table(logCPM.obs, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-obs_DecontX.csv", sep = ",")
# write.table(logCPM.fit, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-fit_DecontX.csv", sep = ",")

```

# 3. Examine significant genes
## 3-1. Compute clusters

```{r}
# Clustering based on fitted values
normalized_counts <- logCPM.obs

# Extract normalized counts for only the significant genes
sig_norm <- data.frame(normalized_counts) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% rownames(tab)[1:sig_num[2]])

colnames(sig_norm) <- c("gene", experiments)
sig_norm <- sig_norm %>% column_to_rownames(var="gene")
head(sig_norm)
```

### Calculate distance
```{r}
library("amap")
gene_dist <- Dist(sig_norm, method="spearman")
gene_hclust <- hclust(gene_dist, method = "ward.D2")
```



### Find the best cluster number k

**Dendrogram**
The height of the dendrogram indicates the order in which the clusters were joined. A more informative dendrogram can be created where the heights reflect the distance between the clusters.
```{r}
plot(gene_hclust, labels=FALSE, hang=-1)
```


**Silhouette coefficient**
```{r}
# Determine the best number of clusters
sil_coef_vector <- c()
for (num_cluster in 2:10){
  sil_cutree <- cutree(gene_hclust, k = num_cluster)
  sil_cl <- silhouette(sil_cutree ,gene_dist)
  sil_cl_summary <- summary(sil_cl)
  sil_coefficient <- as.vector(sil_cl_summary$si.summary["Mean"])
  sil_coef_vector<- c(sil_coef_vector, sil_coefficient)
}
#summary(sil_cl)
```

```{r}
plot(2:10,sil_coef_vector,"b", xlab="k", ylab="silhouette coefficient")
```

**C-index**
```{r}
# library("clusterSim")
# # Determine the best number of clusters
# c_index_vector <- c()
# for (num_cluster in 3:20){
#   c_cutree <- cutree(gene_hclust, k = num_cluster)
#   c_cl <- index.C(gene_dist, c_cutree)
#   c_index_vector<- c(c_index_vector, c_cl)
# }
```

```{r}
#plot(3:20,c_index_vector,"b", xlab="k", ylab="C-index")
```

### Clustering
```{r}
gene_cluster <- data.frame(cutree(gene_hclust, k = 3)) %>%
  rownames_to_column()
colnames(gene_cluster) <- c("gene", "cluster_number")
head(gene_cluster)
```

### Save cluster number as csv
```{r}
cluster_number_merged <- merge(x = tab[1:sig_num[2],], y = gene_cluster, by.x = 0 , by.y = "gene", all = TRUE)
```

```{r}
# write.table(cluster_number_merged, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_pseudobulk_cluster_DecontX_obs.csv", sep = ",")
```


## 3-2. Plot clusters

### 1) General cluster shape
```{r}
histogram <- c()
for(i in 1:3) {

  cluster_genes <- gene_cluster[gene_cluster$cluster_number == i, "gene"]
  logCPM.fit.i <- data.frame(logCPM.obs[cluster_genes,])
  colnames(logCPM.fit.i) <- hours
  histogram <- c(histogram, length(cluster_genes))
  
  
  # Commented line below shows trend lind of each genes
  # plot(hours, logCPM.fit.i[1,], type = "l", ylab="log-CPM", main=paste0("cluster ", i, ": ", length(cluster_genes), " genes"), pch=16)
  # for (j in 2:10){#length(cluster_genes)) {
  #    lines(hours, logCPM.fit.i[j,])
  # }
  
  # Box plot of clusters through timepoint
  box <- boxplot(logCPM.fit.i, ylim=c(-15, 15), ylab="log-CPM", main=paste0("cluster ", i, ": ", length(cluster_genes), " genes"), plot = FALSE)
  boxaverage <- data.frame(average = box$stats[3,], upper_CI = box$conf[2,], lower_CI = box$conf[1,], time = hours)

  cluster_exp_plot <- ggplot(boxaverage) +
    geom_errorbar(aes(hours, ymin = lower_CI, ymax = upper_CI), width = 0.2) +
    geom_line(aes(x = time, y = average), size = 2) +
    geom_point(aes(x = time, y = average)) +
    scale_x_continuous(breaks = hours) +
    labs(title = paste0("Cluster ", i, ": ", length(cluster_genes), " genes")) +
    theme_bw() +
    theme(plot.title = element_text(size=15, hjust=0.5))
  print(cluster_exp_plot)
  
}


#histogram <- as.dataframe(histogram)
```


```{r}
# plot(histogram, type="h", lwd=10, ylim=c(0, 6000), xlab="Cluster", ylab="Cluster Size")
# text(histogram, format(histogram), pos=3, col="red", cex=0.8)
```


### 2) Plot average trend per cell type
```{r}
sig_genes_celltype <- data.frame("gene" = rownames(sig_norm), str_split(rownames(sig_norm), "_", simplify=TRUE))

for(celltype in celltype_OI) {

  cluster_genes <- sig_genes_celltype$gene[sig_genes_celltype$X2 == celltype]
  logCPM.fit.i <- data.frame(logCPM.obs[cluster_genes,])
  colnames(logCPM.fit.i) <- hours

  # Box plot of clusters through timepoint
  box <- boxplot(logCPM.fit.i, ylim=c(-15, 15), ylab="log-CPM", main=paste0("cluster ", i, ": ", length(cluster_genes), " genes"), plot = FALSE)
  boxaverage <- data.frame(average = box$stats[3,], upper_CI = box$conf[2,], lower_CI = box$conf[1,], time = hours)

  cluster_exp_plot <- ggplot(boxaverage) +
    geom_errorbar(aes(hours, ymin = lower_CI, ymax = upper_CI), width = 0.2) +
    geom_line(aes(x = time, y = average), size = 2) +
    geom_point(aes(x = time, y = average)) +
    scale_x_continuous(breaks = hours) +
    labs(title = paste0("Average trend of significant ", celltype, ": ", length(cluster_genes), " genes")) +
    theme_bw() +
    theme(plot.title = element_text(size=15, hjust=0.5))
  print(cluster_exp_plot)
  
}


#histogram <- as.dataframe(histogram)
```

### 3) Number of DE genes per cell type
```{r}
plot_numDE_celltype_df <- data.frame(table(sig_genes_celltype$X2))

plot_numDE_celltype <- ggplot(plot_numDE_celltype_df, aes(x=Var1, y = Freq)) +
  geom_col() +
  geom_text(aes(y = (Freq + 200), label = Freq)) +
  labs(x = "Cell Type", y = "Number of DE genes") +
  theme_bw()
                                
plot_numDE_celltype
```


### 4) Cell type proportion in a cluster
```{r}
# Cluster info with gene symbol & cell type information
gene_celltype_matrix <- str_split(cluster_number_merged$Row.names, "_", simplify=TRUE)
colnames(gene_celltype_matrix) <- c("symbol", "celltype")
cluster_number_merged_with_gene_celltype <- cbind(cluster_number_merged, gene_celltype_matrix)

for (i in 1:length(levels(factor(cluster_number_merged_with_gene_celltype$cluster_number)))){
  cluster_celltype <- cluster_number_merged_with_gene_celltype[cluster_number_merged_with_gene_celltype$cluster_number == i,"celltype"] %>% table()
  cluster_celltype_df <- data.frame(celltype = names(cluster_celltype), freq = as.numeric(cluster_celltype))
  
  cluster_celltype_df <- cluster_celltype_df %>% 
    arrange(desc(celltype)) %>%
    mutate(prop = freq / sum(cluster_celltype_df$freq) *100) %>%
    mutate(ypos = cumsum(prop)- 0.5*prop )
  
  ggplot_celltype_proportion <- ggplot(cluster_celltype_df, aes(x="", y=prop, fill=celltype)) +
    geom_bar(stat="identity", width=1, color = "white") +
    coord_polar("y", start=0) +
    theme_void() +
    geom_text(aes(y = ypos, label = freq, fontface = 2), color = "white", size=4) +
    labs(title = paste0("Celltype Propotion of Cluster ", i))
  
  print(ggplot_celltype_proportion)
}
```

### 5) Heatmap

Heatmap
- explore clustering of the significant genes

```{r}
normalized_counts <- logCPM.obs

# Extract normalized counts for only the significant genes
sig_norm_heatmap <- data.frame(normalized_counts) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% rownames(tab)[1:20])

colnames(sig_norm_heatmap) <- c("gene", experiments)

sig_norm_heatmap <- column_to_rownames(sig_norm_heatmap, var = "gene")

# # Set up row names
# gene_symbol <- c()
# for (n in 1:length(sig_norm_heatmap$gene)) {
#   gene_symbol[n] <- tab$Symbol_celltype[which(rownames(tab) == sig_norm_heatmap$gene[n])]
# }
# 
# rownames(sig_norm_heatmap) <- gene_symbol
```

```{r}
# Set a color palette
heat_colors <- brewer.pal(20, "RdYlBu")

# Run pheatmap using the metadata for the annotation
pheatmap(sig_norm_heatmap[, 1:length(colnames(sig_norm_heatmap))],
         color = heat_colors,
         cluster_rows = T,
         cluster_cols = F, # Make this TRUE if you want to look cluster between different time points
         show_rownames = T,
         border_color = NA,
         fontsize = 10,
         scale = "row",
         fontsize_row = 10,
         height = 20,
         main = paste0("Top DE gene clusters"))
```



## 3-3. Common genes across cell types
```{r}
# Create false dataframe
common_genes <- as.data.frame(matrix(data = FALSE, nrow = length(levels(factor(cluster_number_merged_with_gene_celltype$symbol))), ncol=7))
colnames(common_genes) <- celltype_OI
rownames(common_genes) <- as.vector(levels(factor(cluster_number_merged_with_gene_celltype$symbol)))

# Change False into True if it exists in 
for (n in 1:dim(cluster_number_merged_with_gene_celltype)[1]) {
  common_genes[cluster_number_merged_with_gene_celltype$symbol[n], cluster_number_merged_with_gene_celltype$celltype[n]] <- TRUE
}

common_genes["count"] <- rowSums(common_genes)

head(common_genes)
```


```{r}
common_genes[common_genes$count == 7, "symbol"]
```

