---
title: "DE_PerTime"
author: "Heesoo Song"
date: '2022 6 15 '
output: html_document
---

```{r, message = FALSE}
library(SeuratObject)
library(Seurat)
library(nichenetr)
library(tidyverse)
library(ggplot2)
library(edgeR)
library(rmarkdown)
library(SingleCellExperiment)
library(DEsingle)
library(Matrix.utils)
```

# 0. Load Dataset
```{r}
IDtoSymbol_matrix <- read.csv("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/IDtoSymbol_matrix.csv")
```

```{r}
# Load processed expression data of interacting cells
hepatectomy_expression = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/seurat_obj_phx_25102021.rds")
```

```{r}
celltype_OI <- c("cholangiocyte","endothelial","hepatocyte","kupffer","monocyte","neutrophil","stellate")

# experiments <- c("0_phase_1", "3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_1", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")
experiments <- c("3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")

#hours <- c(0, 3, 6, 12, 24, 36, 48, 48, 72, 96, 120, 168)
hours <- c(3, 6, 12, 24, 36, 48, 72, 96, 120, 168)
```

```{r}
hepa_celltypeOI = subset(x = hepatectomy_expression, idents = celltype_OI)

hepa_celltypeOI@meta.data$time_experiment <- factor(hepa_celltypeOI@meta.data$time_experiment, levels = experiments)
```

```{r}
rm(hepatectomy_expression)
```

## 0-1. Change ENSEMBL to gene symbols
```{r}
tp1 <- "3_phase_2"
tp2 <- "6_phase_2"

hepa_timeOI <- subset(x = hepa_celltypeOI, subset = time_experiment %in% c(tp1, tp2))

sce <- as.SingleCellExperiment(hepa_timeOI)
```

```{r}
counts(sce)[1:6,1:4]
```

```{r}
## change ENSEMBL to gene symbol in column names

# library(biomaRt)
# ensembl <- useEnsembl(biomart="ensembl", dataset = "mmusculus_gene_ensembl")
# IDtoSymbol_matrix <- getBM(attributes=c("ensembl_gene_id","external_gene_name"), filters = "ensembl_gene_id", values = unique(colnames(expression_tp1)), mart = ensembl)

IDtoSymbol_matrix <- read.csv("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/IDtoSymbol_matrix.csv")

head(IDtoSymbol_matrix)
```

```{r}
rownames_sce <- data.frame("ENSEMBL" = rownames(sce))
rownames_sce["order"] <- c(1:dim(rownames_sce)[1])
gene_ID <- merge(rownames_sce, IDtoSymbol_matrix, by.x="ENSEMBL", by.y = "ensembl_gene_id", all.x=TRUE)

# Fill rest of NA with ENSEMBL ID (ID no longer in the ENSEMBL database)
gene_ID$external_gene_name[which(gene_ID$external_gene_name == "")] <- gene_ID$ENSEMBL[which(gene_ID$external_gene_name == "")]
gene_ID$external_gene_name[which(is.na(gene_ID$external_gene_name))] <- gene_ID$ENSEMBL[which(is.na(gene_ID$external_gene_name))]
gene_ID <- gene_ID[order(gene_ID$order),]

# Change rownames of SCE
rownames(sce) <- gene_ID$external_gene_name

counts(sce)[1:6,1:4]
```

# 1. DE analysis (DEsingle)

```{r}
library(tictoc)
```

## 1-1. DEsingle
```{r}
#group <- factor(sce$time_experiment)

# library(BiocParallel)
# param <- SnowParam(workers = 8, type = "SOCK", progressbar = TRUE)
# register(param)
# 
# tic("DEsingle")
# results <- DEsingle(counts = sce, group = group, parallel = TRUE, BPPARAM = param)
# toc()
# 
# results.classified <- DEtype(results = results, threshold = 0.05)
```

## 1-2. EdgeR (Pseudobulk)
If we make a pseudobulk per cell type cluster, it will end up with high false positive rate. Effect size (logFC)/ SE (standard error) when calculating p-values. However in logFC, pseudobulk 
```{r}
# # Create aggregate matrix
# group <- colData(sce)[, c("ident", "time_experiment")]
# 
# pb <- aggregate.Matrix(t(counts(sce)),
#                        groupings = group, fun = "sum")
# 
# 
# dim(pb)
# pb[1:5, 1:6]
```

```{r}
# y <- DGEList(counts = t(pb))
```

```{r}
# # filter out lowly expressed genes
# keep <- filterByExpr(y)
# y <- y[keep, , keep.lib.sizes = FALSE]
# 
# # TMM Normalisation
# y <- calcNormFactors(y)
# y$samples
```
```{r}
# plotMDS(y)
```

```{r}
# timepoint <- factor(sapply(stringr::str_split(rownames(pb), pattern="_", n=2), '[', 2))
# cluster <- factor(sapply(stringr::str_split(rownames(pb), pattern="_", n=2), '[', 1))
# 
# design <- model.matrix(~timepoint+cluster)
# design
```

```{r}
# y <- estimateDisp(y, design)
# sqrt(y$common.dispersion)
# 
# # plot the tagwise dispersion against log2-CPM (counts per million)
# plotBCV(y)
```

```{r}
# fit <- glmFit(y, design, robust=TRUE)
```

```{r}
# lrt <- glmLRT(fit, coef = 2)
# topTags(lrt)
```

```{r}
# o <- order(lrt$table$PValue)
# y$counts[o[1:10],1:4]
# # Differences between tp1 vs. tp2 can be observed in each cell type
```

The total number of differentially expressed genes at 5% FDR
```{r}
# sig_num <- summary(decideTests(lrt))
# sig_num
```

```{r}
# plotMD(lrt)
# abline(h=c(-1, 1), col="blue") # Indicates 2 fold changes
```

```{r}
# go <- goana(lrt)
# topGO(go, ont="BP", sort="Up", n=30, truncate=30)
# Error in goana.default(de = DEGenes, universe = universe, ...) : No annotated genes found in universe
```

```{r}
# Genes with most significant time effects
# H0: no trend/pattern in expression level over time
# tab <- as.data.frame(topTags(lrt, n=dim(lrt)[1]))
# head(tab)
```

```{r}
# logCPM.obs <- edgeR::cpm(y, log=TRUE, prior.count=lrt$prior.count)
# logCPM.fit <- edgeR::cpm(lrt, log=TRUE)
```

```{r}
# write.table(logCPM.obs, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-obs_DecontX.csv", sep = ",")
# write.table(logCPM.fit, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-fit_DecontX.csv", sep = ",")

```

## 1-3. Seurat FindMarkers

Compare differentially expressed genes between two timepoints in a cell type.
adjusted p=1 means inability to reject the null at any significance level; in intuitive terms, this means ``no evidence against the null''
```{r}
DE_list <- list()
for (celltype in celltype_OI){
  onecelltype_seurat <- subset(hepa_timeOI, idents = celltype)
  DE_genes <- FindMarkers(onecelltype_seurat, ident.1 = tp2, ident.2 = tp1, group.by = "time_experiment", logfc.threshold = 0.5)
  
  # add gene symbol
  DE_genes <- merge(DE_genes, IDtoSymbol_matrix, by.x = 0, by.y = "ensembl_gene_id", all.x = TRUE) %>% column_to_rownames("Row.names")
  DE_genes <- DE_genes[DE_genes$external_gene_name != "",]
  DE_genes <- DE_genes[!is.na(DE_genes$external_gene_name),]
  DE_genes <- DE_genes[DE_genes$p_val_adj < 0.5,] # remove adjusted p-value < 0.5
  DE_genes <- DE_genes[order(DE_genes$p_val),]
  
  # add to list
  DE_list[[celltype]] <- DE_genes
}
```

# 2. 



