---
title: "DE_pseudobulk_clusterPerCelltype"
author: "Heesoo Song"
date: '2022 7 24 '
output: html_document
---

Extension of *DE_EdgeR.Rmd*. Here, we will create aggregate matrix to observe change spanning over celltype rather than time point.
https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf

```{r}
#install.packages('Matrix.utils')
#install.packages("statmod")
#BiocManager::install("org.Mm.eg.db")
```


```{r, message = FALSE}
library(SeuratObject)
library(Seurat)
library(nichenetr)
library(edgeR)
library(tidyverse)
library("ggplot2")
library(muscat)
library(SingleCellExperiment)
library(purrr)
library(Matrix.utils)
library(RColorBrewer)
library(pheatmap)
library(cluster)
library(celda)
```

```{r}
# Load processed expression data of interacting cells
hepatectomy_expression = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/seurat_obj_phx_25102021.rds")

# Total 91021 cells, with 5000 genes
```

```{r}
celltype_OI <- c("cholangiocyte","endothelial","hepatocyte","kupffer","monocyte","neutrophil","stellate")
experiments <- c("3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")

hours <- c(3, 6, 12, 24, 36, 48, 72, 96, 120, 168)
```

```{r}
hepa_celltypeOI = subset(x = hepatectomy_expression, idents = celltype_OI)

hepa_celltypeOI@meta.data$time_experiment <- factor(hepa_celltypeOI@meta.data$time_experiment, levels = experiments)

hepa_celltypeOI = subset(x = hepa_celltypeOI, subset = time_experiment %in% experiments)
```

```{r}
rm(hepatectomy_expression)
```


```{r}
# Load decontaminated count matrix
sce <- readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/sce_DecontX.rds")
```

## Create aggregate matrix
```{r}
# Identify groups for aggregation of counts
#groups <- colData(sce)[, c("cluster_id", "time_experiment")]
groups <- colData(sce)[, c("ident", "time_experiment")]
```

```{r}
# Create aggregate matrix
# pb <- aggregate.Matrix(t(counts(sce)),
#                        groupings = groups, fun = "sum")
pb <- aggregate.Matrix(t(decontXcounts(sce)),
                       groupings = groups, fun = "sum")

dim(pb)
pb[1:6, 1:6]
```

```{r}
# Not every cluster is present in all samples; create a vector that represents how to split samples
# But actually we do have all cell types! :) Maybe skip this step
splitf <- sapply(stringr::str_split(rownames(pb), pattern="_", n=2), '[', 1)
```

```{r}
# Turn into a list and split the list into components for each cluster and transform, so rows are genes and columns are samples and make rownames as the sample IDs
library(magrittr)
pb <- split.data.frame(pb, 
                       factor(splitf)) %>%
  lapply(function(u) 
    set_colnames(t(u), 
                 stringr::str_extract(rownames(u), "(?<=_)[:alnum:]+_phase_[0-9]"))) 


# Explore the different components of list
names(pb)
```

```{r}
pb[[1]][1:6, 1:6]
```

```{r}
#Change row names into geneID + celltype
for (i in 1:length(names(pb))){
  rownames(pb[[i]]) <- paste0(rownames(pb[[i]]), "_", names(pb)[i])
}

pb[[1]][1:6, 1:6]
```


```{r}
# Combine matrices
pb_combined <- pb[[1]]
dim(pb_combined)

for (i in 2:length(names(pb))) {
  pb_combined <- rbind(pb_combined, pb[[i]])
}
dim(pb_combined)
cluster_counts <- as.data.frame(as.matrix(pb_combined))
```


```{r}
# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$ident, sce$time_experiment)
```

# 2. Differential gene expression
The aim is to examine the relationship between gene abundances and time points per each cell-type. In other words, we are looking for genes that change expression level over time in a time course experiment. Given enough number of time points, it is reasonable to assume smooth changes in gene expression. In this way, polynomial or a cubic spline curve can be used to model gene expression along time with a certain number of degrees of freedom.


Perform DE analysis

Explore aggregated count data by MDS (multi-dimensional scaling) plots. This visualizes the differences between the expression profiles of different samples (cholangiocyte) in two dimensions.

```{r}
plotMDS(cluster_counts, labels=colnames(cluster_counts))
```

**PCA per cell type**
```{r}
for (i in 1:length(pb)){
  colnames(pb[[i]]) <- hours
}

for (i in 1:length(pb)){
  celltypeMDS <- plotMDS(pb[[i]], labels=colnames(pb[[i]]), main = names(pb[i]))
  print(celltypeMDS)
}
```

Use cubic regression spline curve with 4 degrees of freedom. The advantage of using a cubic spline curve is that it provides more stable fit at the end points compared to a polynomial. The spline curve with 3 degrees of freedom has 2 knots where cucbic polynomicals are splined together.
```{r}
library(splines)
X <- ns(hours, df = 4)
design <- model.matrix(~X)
design
```
Estimate dispersion. The negative binomial (NB) dispersion is estimated. This returns the **DGEList** object with additional entries for the estimated NB dispersion for each gene.
```{r}
y <- DGEList(counts = cluster_counts, group = hours)
```

```{r}
# filter out lowly expressed genes
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes = FALSE]

# TMM Normalisation
# y <- calcNormFactors(y)
# y$samples
```


```{r}
y <- estimateDisp(y, design)
sqrt(y$common.dispersion)

# plot the tagwise dispersion against log2-CPM (counts per million)
plotBCV(y)
```


Fit a quasi-likelihood NB generalized log-linear model

```{r}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```

Fit a quasi-likelihood negative binomial generalized log-linear model to count data.
```{r}
fit <- glmQLFTest(fit, coef=2:5)
```

Multiple testing across genes and contrasts. The total number of genes with significant (5% FDR) changes at different time points. 
All spline coefficients are tested together. Thus t-tests for the individual coefficients and the logFC values do not have any interpretable meaning.

```{r}
sig_num <- summary(decideTests(fit))
sig_num
```

```{r}
# Genes with most significant time effects
# H0: no trend/pattern in expression level over time
tab <- as.data.frame(topTags(fit, n=dim(fit)[1]))
head(tab)
```

LogFC = log fold change, log difference between groups.
LogCPM = log count per million, measure of expression level.


```{r}
logCPM.obs <- edgeR::cpm(y, log=TRUE, prior.count=fit$prior.count)
logCPM.fit <- edgeR::cpm(fit, log=TRUE)
```

```{r}
# write.table(logCPM.obs, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-obs_DecontX.csv", sep = ",")
# write.table(logCPM.fit, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_geneID+celltype_logCPM-fit_DecontX.csv", sep = ",")

```

# 3. Examine significant genes
## 3-1. Compute clusters per cell type

```{r}
# Clustering based on observed values
normalized_counts.obs <- logCPM.obs

# Extract normalized counts for only the significant genes
sig_norm.obs <- data.frame(normalized_counts.obs) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% rownames(tab)[1:sig_num[2]])

colnames(sig_norm.obs) <- c("gene", experiments)
sig_norm.obs <- sig_norm.obs %>% column_to_rownames(var="gene")

sig_norm.obs$celltype <- str_split(rownames(sig_norm.obs), "_", simplify=TRUE)[,2]
head(sig_norm.obs)

########################################
# Clustering based on fitted values
normalized_counts.fit <- logCPM.fit

# Extract normalized counts for only the significant genes
sig_norm.fit <- data.frame(normalized_counts.fit) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% rownames(tab)[1:sig_num[2]])

colnames(sig_norm.fit) <- c("gene", experiments)
sig_norm.fit <- sig_norm.fit %>% column_to_rownames(var="gene")

sig_norm.fit$celltype <- str_split(rownames(sig_norm.fit), "_", simplify=TRUE)[,2]
```

### Calculate distance
```{r}
# Filter sig_norm with one cell type
celltype = celltype_OI[7]

sig_norm_onecelltype.obs <- sig_norm.obs[sig_norm.obs$celltype == celltype,]

library("amap")
gene_dist.obs <- Dist(sig_norm_onecelltype.obs[,experiments], method="spearman")
gene_hclust.obs <- hclust(gene_dist.obs, method = "ward.D2")

######################################
sig_norm_onecelltype.fit <- sig_norm.fit[sig_norm.fit$celltype == celltype,]

gene_dist.fit <- Dist(sig_norm_onecelltype.fit[,experiments], method="spearman")
gene_hclust.fit <- hclust(gene_dist.fit, method = "ward.D2")
```



### Find the best cluster number k

**Dendrogram**
The height of the dendrogram indicates the order in which the clusters were joined. A more informative dendrogram can be created where the heights reflect the distance between the clusters.
```{r}
plot(gene_hclust.obs, labels=FALSE, hang=-1, main = paste0("Cluster Dendrogram : ", celltype, " (obs)"))
plot(gene_hclust.fit, labels=FALSE, hang=-1, main = paste0("Cluster Dendrogram : ", celltype, " (fit)"))
```


**Silhouette coefficient**
```{r}
# Determine the best number of clusters
sil_coef_vector.obs <- c()
for (num_cluster in 2:10){
  sil_cutree.obs <- cutree(gene_hclust.obs, k = num_cluster)
  sil_cl.obs <- silhouette(sil_cutree.obs ,gene_dist.obs)
  sil_cl_summary.obs <- summary(sil_cl.obs)
  sil_coefficient.obs <- as.vector(sil_cl_summary.obs$si.summary["Mean"])
  sil_coef_vector.obs <- c(sil_coef_vector.obs, sil_coefficient.obs)
}

#################################
# Determine the best number of clusters
sil_coef_vector.fit <- c()
for (num_cluster in 2:10){
  sil_cutree.fit <- cutree(gene_hclust.fit, k = num_cluster)
  sil_cl.fit <- silhouette(sil_cutree.fit ,gene_dist.fit)
  sil_cl_summary.fit <- summary(sil_cl.fit)
  sil_coefficient.fit <- as.vector(sil_cl_summary.fit$si.summary["Mean"])
  sil_coef_vector.fit <- c(sil_coef_vector.fit, sil_coefficient.fit)
}
```

```{r}
plot(2:10,sil_coef_vector.obs,"b", xlab="k", ylab="silhouette coefficient", main = "logCPM.obs")
plot(2:10,sil_coef_vector.fit,"b", xlab="k", ylab="silhouette coefficient", main = "logCPM.fit")
```


### Clustering
```{r}
gene_cluster.obs <- data.frame(cutree(gene_hclust.obs, k = 2)) %>%
  rownames_to_column()
colnames(gene_cluster.obs) <- c("gene", "cluster_number")
head(gene_cluster.obs)

######################################
gene_cluster.fit <- data.frame(cutree(gene_hclust.fit, k = 2)) %>%
  rownames_to_column()
colnames(gene_cluster.fit) <- c("gene", "cluster_number")
```

### Save cluster number as csv
```{r}
# cluster_number_merged <- merge(x = tab[1:sig_num[2],], y = gene_cluster, by.x = 0 , by.y = "gene", all = TRUE)

# write.table(cluster_number_merged, file="C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/DE_pseudobulk_cluster_DecontX_obs.csv", sep = ",")
```

### General cluster shape (obs)
```{r}
#histogram <- c()
for(i in 1:2) {

  cluster_genes.obs <- gene_cluster.obs[gene_cluster.obs$cluster_number == i, "gene"]
  logCPM.obs.i <- data.frame(logCPM.obs[cluster_genes.obs,])
  colnames(logCPM.obs.i) <- hours
  #histogram <- c(histogram, length(cluster_genes))
  
  # Box plot of clusters through timepoint
  box.obs <- boxplot(logCPM.obs.i, ylim=c(-15, 15), ylab="log-CPM", main=paste0("cluster ", i, ": ", length(cluster_genes.obs), " genes"), plot = FALSE)
  boxaverage.obs <- data.frame(average = box.obs$stats[3,], upper_CI = box.obs$conf[2,], lower_CI = box.obs$conf[1,], time = hours)

  cluster_exp_plot.obs <- ggplot(boxaverage.obs) +
    geom_errorbar(aes(hours, ymin = lower_CI, ymax = upper_CI), width = 0.2) +
    geom_line(aes(x = time, y = average), size = 2) +
    geom_point(aes(x = time, y = average)) +
    scale_x_continuous(breaks = hours) +
    labs(title = paste0("Cluter ", i, "(obs) : ", length(cluster_genes.obs), " genes")) +
    theme_bw() +
    theme(plot.title = element_text(size=15, hjust=0.5))
  print(cluster_exp_plot.obs)
  
}

#histogram <- as.dataframe(histogram)
```
### General cluster shape (fit)
```{r}
#histogram <- c()
for(i in 1:2) {

  cluster_genes.fit <- gene_cluster.fit[gene_cluster.fit$cluster_number == i, "gene"]
  logCPM.fit.i <- data.frame(logCPM.fit[cluster_genes.fit,])
  colnames(logCPM.fit.i) <- hours
  #histogram <- c(histogram, length(cluster_genes))
  
  # Box plot of clusters through timepoint
  box.fit <- boxplot(logCPM.fit.i, ylim=c(-15, 15), ylab="log-CPM", main=paste0("cluster ", i, ": ", length(cluster_genes.fit), " genes"), plot = FALSE)
  boxaverage.fit <- data.frame(average = box.fit$stats[3,], upper_CI = box.fit$conf[2,], lower_CI = box.fit$conf[1,], time = hours)

  cluster_exp_plot.fit <- ggplot(boxaverage.fit) +
    geom_errorbar(aes(hours, ymin = lower_CI, ymax = upper_CI), width = 0.2) +
    geom_line(aes(x = time, y = average), size = 2) +
    geom_point(aes(x = time, y = average)) +
    scale_x_continuous(breaks = hours) +
    labs(title = paste0("Cluter ", i, "(fit) : ", length(cluster_genes.fit), " genes")) +
    theme_bw() +
    theme(plot.title = element_text(size=15, hjust=0.5))
  print(cluster_exp_plot.fit)
  
}

#histogram <- as.dataframe(histogram)
```

### Overlapping genes between clustering with observed counts and fitted counts
```{r}
library(ggvenn)

for (i in 1:2){
  cluster_genes.obs <- gene_cluster.obs[gene_cluster.obs$cluster_number == i, "gene"]
  cluster_genes.fit <- gene_cluster.fit[gene_cluster.fit$cluster_number == i, "gene"]
  
  venn <- list(obs = cluster_genes.obs, fit = cluster_genes.fit)
  
  diagram <- ggvenn(
    venn, 
    fill_color = c("#0073C2FF", "#EFC000FF"),
    stroke_size = 0.5, set_name_size = 4
    ) + ggtitle(paste0("Cluster ", i))
  print(diagram)
}
```

