---
title: "Dynamic_NicheNet_PerTime"
author: "Heesoo Song"
date: '2022 6 25 '
output: html_document
---


# 0. Install/Load packages

```{r}
#install.packages("devtools")
#install.packages("htmltools")
#BiocManager::install("limma")
#devtools::install_github("saeyslab/nichenetr")
#install.packages("tidyverse")
```


```{r message=FALSE, warning=FALSE}
library(nichenetr)
library(tidyverse)
library(Matrix)
library(dplyr)
library(rmarkdown)
library(tibble)
library(Seurat)
```

# 1. Load datasets

## 1) NicheNet prior models
Numbers in the ligand-target matrix represents the ligand-target regulatory potential scores between all pairs of ligands and target genes. These scores are calculated using network propagation methods on the integrated networks to propagate the signal from a ligand, over receptors, signaling proteins and transcriptional regulators, to end at target genes.
```{r cache=TRUE}
# Load ligand-target prior model
#ligand_target_matrix = readRDS(url("https://zenodo.org/record/5884439/files/ligand_target_matrix_nsga2r_final_mouse.rds"))  # New version
#ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))  # Old version

# in case bad internet connection
ligand_target_matrix = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/ligand_target_matrix_nsga2r_final_mouse.rds")
ligand_target_matrix[1:5, 1:5]
```
Putative ligand-receptor links were gathered from NicheNet's ligand-receptor data sources.
```{r cache=TRUE}
#lr_network = readRDS(url("https://zenodo.org/record/5884439/files/lr_network_mouse_21122021.rds")) # new version
#lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds")) # old version

# in case bad internet connection
lr_network = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/lr_network_mouse_21122021.rds")
head(lr_network)
```

```{r cache=TRUE}
# get the weights of the ligand-receptor interactions as used in the NicheNet model
#weighted_networks = readRDS(url("https://zenodo.org/record/5884439/files/weighted_networks_nsga2r_final_mouse.rds")) # New version
#weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds")) # Old version

# in case internet connection is bad
weighted_networks = readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/weighted_networks_nsga2r_final_mouse.rds")
weighted_networks_lr = weighted_networks$lr_sig %>% inner_join(lr_network %>% distinct(from,to), by = c("from","to"))
head(weighted_networks$lr_sig)
head(weighted_networks$gr)
```


## 2) logFC & significance of all genes
```{r cache=TRUE}
summary_df <- readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/Summary_allgenes_PerTime.rds")
paged_table(head(summary_df))
```

## 3) Expression Data
```{r}
hepa_rds <- readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/seurat_obj_phx_25102021.rds")
sce <- readRDS("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/sce_DecontX.rds")
```

```{r}
# Preprocess expression data
## Subset to seven cell-types
celltype_OI <- c("cholangiocyte","endothelial","hepatocyte","kupffer","monocyte","neutrophil","stellate")
hepa_celltypeOI <- subset(x = hepa_rds, idents = celltype_OI)

## Factorize experiment timepoint
#experiments <- c("0_phase_1", "3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_1", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")
#hours <- c(0, 3, 6, 12, 24, 36, 48, 48, 72, 96, 120, 168)
experiments <- c("3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")
hours <- c(3, 6, 12, 24, 36, 48, 72, 96, 120, 168)

hepa_celltypeOI@meta.data$time_experiment <- factor(hepa_celltypeOI@meta.data$time_experiment, levels = experiments)
hepa_celltypeOI = subset(x = hepa_celltypeOI, subset = time_experiment %in% experiments)

rm(hepa_rds)
```

```{r}
library(celda)
hepa_sample_info <- colData(sce)
hepa_expression <- t(decontXcounts(sce))
```

```{r}
## change ENSEMBL to gene symbol in column names

# library(biomaRt)
# ensembl <- useEnsembl(biomart="ensembl", dataset = "mmusculus_gene_ensembl")
# IDtoSymbol_matrix <- getBM(attributes=c("ensembl_gene_id","external_gene_name"), filters = "ensembl_gene_id", values = unique(colnames(hepa_expression)), mart = ensembl)

IDtoSymbol_matrix <- read.csv("C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/IDtoSymbol_matrix.csv")
```

```{r}
expression_colnames <- data.frame(colnames(hepa_expression))
expression_colnames["order"] <- c(1:dim(expression_colnames)[1])
symbol_list <- merge(expression_colnames, IDtoSymbol_matrix,by.x = "colnames.hepa_expression.", by.y = "ensembl_gene_id", all.x = TRUE)

## Fill rest of NA with ENSEMBL ID (ID no longer in the ENSEMBL database)
symbol_list$external_gene_name[is.na(symbol_list$external_gene_name)] <- symbol_list$colnames.hepa_expression.[is.na(symbol_list$external_gene_name)]
symbol_list <- symbol_list[order(symbol_list$order),]

## In old version, the gene names are in uppercase characters, while the new version is not.
### New version
colnames(hepa_expression) <- symbol_list$external_gene_name

### Old version
#colnames(hepa_expression) <- casefold(symbol_list$external_gene_name, upper=TRUE)
#hepa_cluster$Symbol <- casefold(hepa_cluster$Symbol, upper=TRUE)

hepa_expression[1:5, 1:5]
```


# 2. Define expressed genes

- sender cell type, receiver cell type. We will analyze 7x7x3 different combinations
- 7 cell types ("cholangiocyte","endothelial","hepatocyte","kupffer","monocyte","neutrophil","stellate")
- 11 time points ("0_phase_1", "3_phase_2", "6_phase_2", "12_phase_2", "24_phase_2", "36_phase_2", "48_phase_1", "48_phase_2", "72_phase_2", "96_phase_2", "120_phase_2", "168_phase_2")
- consider a gene to be expressed when it is expressed in at least 10% of cells
- filter cells by each time point (sample) rather than to filter by cell type with all time points as the genes may be expressed in only at few time points while lowly expressed in the other time points.

```{r}
main_path <- "C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/Dynamic_NicheNet/Dynamic_NicheNet_PerTime"
sender_celltype = "all"

ligand_target_list <- list()

for (receiver_celltype in celltype_OI){

  for (i in 1:(length(hours)-1)){
    sender_time = as.character(hours)[i]
    receiver_time = as.character(hours)[i+1]

      
    print(paste0("Start Running: ", sender_celltype, sender_time, "_", receiver_celltype, receiver_time))
    
    
    # Extract sender & receiver cells
    if (sender_celltype == "all"){
      expressed_genes_sender <- c()
      for (celltype in celltype_OI){
        # Define sender
        sender_ids <- row.names(hepa_sample_info[which(hepa_sample_info$celltype == celltype & hepa_sample_info$time == sender_time),])
        expressed_genes_sender_onetype <- colSums(hepa_expression[sender_ids,] >= 1) %>% sapply(function(x){x/length(sender_ids)*100}) %>% .[. >= 10] %>% names()
        expressed_genes_sender <- c(expressed_genes_sender, expressed_genes_sender_onetype)
      }
      # Define receiver
      receiver_ids <- row.names(hepa_sample_info[which(hepa_sample_info$celltype == receiver_celltype & hepa_sample_info$time == receiver_time),])
      expressed_genes_receiver <- colSums(hepa_expression[receiver_ids,] >= 1) %>% sapply(function(x){x/length(receiver_ids)*100}) %>% .[. >= 10] %>% names()
    
      expressed_genes_sender <- expressed_genes_sender %>% unique()
      expressed_genes_receiver <- expressed_genes_receiver %>% unique()
    
      } else{
    
        # Define sender
        sender_ids <- row.names(hepa_sample_info[which(hepa_sample_info$celltype == sender_celltype & hepa_sample_info$time == sender_time),])
        expressed_genes_sender <- colSums(hepa_expression[sender_ids,] >= 1) %>% sapply(function(x){x/length(sender_ids)*100}) %>% .[. >= 10] %>% names() %>% unique()
    
        # Define receiver
        receiver_ids <- row.names(hepa_sample_info[which(hepa_sample_info$celltype == receiver_celltype & hepa_sample_info$time == receiver_time),])
        expressed_genes_receiver <- colSums(hepa_expression[receiver_ids,] >= 1) %>% sapply(function(x){x/length(receiver_ids)*100}) %>% .[. >= 10] %>% names() %>% unique()
    }
    
    
    
    
    
    
    # 3. Define gene set of interest and background of genes
    
    geneset_oi <- summary_df$gene[summary_df$celltype == receiver_celltype & summary_df$timepoint == receiver_time & summary_df$p_val < 0.5] %>% .[. %in% rownames(ligand_target_matrix)]

    
    # cluster genes that are not expressed in the receiver cells
    geneset_oi <- geneset_oi[geneset_oi %in% expressed_genes_receiver]
    
    background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
    
    
    
    
    # Avoid error caused by Null gene set of interest
    if (length(geneset_oi) > 10){
      
      dir.create(file.path(main_path, paste0("Dynamic_NicheNet_PerTime_", sender_celltype, sender_time, "_", receiver_celltype, receiver_time)))
    vis_path <- paste0(main_path, "/Dynamic_NicheNet_PerTime_", sender_celltype, sender_time, "_", receiver_celltype, receiver_time, "/")
      
      # 4. Define a set of potential ligands
      
      ligands = lr_network %>% pull(from) %>% unique()
      expressed_ligands = intersect(ligands,expressed_genes_sender)
      
      receptors = lr_network %>% pull(to) %>% unique()
      expressed_receptors = intersect(receptors,expressed_genes_receiver)
      
      lr_network_expressed = lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) 
      paged_table(head(lr_network_expressed))
      
      potential_ligands = lr_network_expressed %>% pull(from) %>% unique() %>% .[. %in% colnames(ligand_target_matrix)]
      
      
      
      
      
      
      
      
      # 5. Perform NicheNet's ligand activity analysis on the gene set of interest
      
      # Prediction (not ranked yet)
      ligand_activities = predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = potential_ligands)
      
      # Rank the prediction
      ligand_activities %>% arrange(-aupr)
      best_upstream_ligands = ligand_activities %>% top_n(20, aupr) %>% arrange(-aupr) %>% pull(test_ligand) %>% unique()
      
      
      pdf(file = paste(vis_path, "Ligand_activity_histogram.pdf", sep=""), width = 12, height = 7)
      # show histogram of ligand activity scores
      p_hist_lig_activity = ggplot(ligand_activities, aes(x=aupr)) + 
        geom_histogram(color="black", fill="darkorange")  + 
        # geom_density(alpha=.1, fill="orange") +
        geom_vline(aes(xintercept=min(ligand_activities %>% top_n(20, aupr) %>% pull(aupr))), color="red", linetype="dashed", size=1) + 
        labs(x="ligand activity (AUPR)", y = "# ligands") +
        theme_classic()
      print(p_hist_lig_activity)
      dev.off()
      
      
      
      
      
      
      
      
      # 6. Infer target genes of top-ranked ligands and visualize in a heatmap
      
      ## 1) infer target genes
      
      active_ligand_target_links_df = best_upstream_ligands %>% lapply(get_weighted_ligand_target_links, geneset = geneset_oi, ligand_target_matrix = ligand_target_matrix, n = 250) %>% bind_rows() %>% drop_na()
      
      paged_table(head(active_ligand_target_links_df))
      
      ## 2) Set cutoff for visualization
      
      # visualization
      active_ligand_target_links = prepare_ligand_target_visualization(ligand_target_df = active_ligand_target_links_df, ligand_target_matrix = ligand_target_matrix, cutoff = 0.25) # score below threshold will be set as 0
      
      ## 3) Plot putatively active ligand-target links
      
      order_ligands = intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
      order_targets = active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links))
      vis_ligand_target = active_ligand_target_links[order_targets,order_ligands] %>% t()
      vis_ligand_target_renamed <- vis_ligand_target
      colnames(vis_ligand_target_renamed) <- make.names(colnames(vis_ligand_target))
      rownames(vis_ligand_target_renamed) <- make.names(rownames(vis_ligand_target))
      
      
      pdf(file = paste(vis_path, "Ligand-target_heatmap.pdf", sep=""), width = 12, height = 7)
      
      p_ligand_target_network = vis_ligand_target_renamed %>% make_heatmap_ggplot("Prioritized ligands",paste0("Predicted target genes (", length(order_targets), ")"), color = "purple",legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential") + scale_fill_gradient2(low = "whitesmoke",  high = "purple", breaks = c(0,0.05,0.10)) + theme(axis.text.x = element_text(face = "italic", size = 6))
      
      print(p_ligand_target_network)
      
      dev.off()
      
      
      capture.output(order_targets, file = paste0(vis_path, "order_targets.txt"))
      
      
      
      
      
      # 7. Validation
      
      ## 1) Ligand-receptor network inference for top-ranked ligands
      
      # get the ligand-receptor network of the top-ranked ligands
      lr_network_top = lr_network %>% filter(from %in% best_upstream_ligands & to %in% expressed_receptors) %>% distinct(from,to)
      best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()
      
      
      lr_network_top_df = weighted_networks$lr_sig %>% filter(from %in% best_upstream_ligands & to %in% best_upstream_receptors)
      
      # convert to a matrix
      lr_network_top_df = lr_network_top_df %>% spread("from","weight",fill = 0)
      lr_network_top_matrix = lr_network_top_df %>% dplyr::select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df$to)
      
      # perform hierarchical clustering to order the ligands and receptors
      # hierarchical clustering groups similar objects into clusters
      # Ward.D2 = Ward algorithm = creates groups such that variance is minimized within clusters. Dissimilarities are squared before clustering.
      dist_receptors = dist(lr_network_top_matrix, method = "binary")
      hclust_receptors = hclust(dist_receptors, method = "ward.D2")
      order_receptors = hclust_receptors$labels[hclust_receptors$order]
      
      dist_ligands = dist(lr_network_top_matrix %>% t(), method = "binary")
      hclust_ligands = hclust(dist_ligands, method = "ward.D2")
      order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]
      
      
      pdf(file = paste(vis_path, "Ligand-receptor_heatmap.pdf", sep=""), width = 12, height = 7)
      
      vis_ligand_receptor_network = lr_network_top_matrix[order_receptors, order_ligands_receptor]
      p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Prioritized ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential")
      print(p_ligand_receptor_network)
      
      dev.off()
      
      
      
      ## 2) Visualize expression of top-predicted ligands and their target genes in a combined heatmap
      
      library(RColorBrewer)
      library(cowplot)
      library(ggpubr)
      
      ###ligand activity matrix**
      
      ligand_aupr_matrix = ligand_activities %>% dplyr::select(aupr) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities$test_ligand)
      
      vis_ligand_aupr = ligand_aupr_matrix[order_ligands, ] %>% as.matrix(ncol = 1) %>% magrittr::set_colnames("AUPR")
      
      
      pdf(file = paste(vis_path, "Ligand_activity.pdf", sep=""), width = 12, height = 7)
      
      p_ligand_aupr = vis_ligand_aupr %>% make_heatmap_ggplot("Prioritized ligands","Ligand activity", color = "darkorange",legend_position = "top", x_axis_position = "top", legend_title = "AUPR\ntarget gene prediction ability)")
      print(p_ligand_aupr)
      
      dev.off()
      
      
      ligand_target_list[[paste0(sender_celltype, sender_time, "_", receiver_celltype, receiver_time)]] <- vis_ligand_target_renamed

      
      # # 8. Temporal Interpretation
      # 
      # 
      # # Extract gene of interest (Example genes for now)
      # ligand <- paste0("Cklf_", celltype_OI)
      # target <- "Uhrf1_hepatocyte"
      # 
      # prep_ggplot <- summary_df[which(summary_df$gene_celltype %in% c(ligand, target)),]
      # head(prep_ggplot)
      # 
      # 
      # ## 1) Line type
      # 
      # ggplot(prep_ggplot, aes(x = timepoint, y = avg_log2FC)) + 
      #   geom_line(aes(group = gene_celltype, colour = gene_celltype, size = LigandTarget, linetype = Significance)) + 
      #   scale_size_manual(values = c(target = 3, ligand = 0.5)) +
      #   scale_linetype_manual(values = c("sig" = "solid", "insig" = "dashed")) +
      #   guides(colour = "none") +
      #   coord_cartesian(clip = "off") +
      #   theme_bw() +
      #   theme(legend.position = "top", plot.margin = margin(0.1, 2.6, 0.1, 0.1, "cm"), plot.title = element_text(hjust = 0.5)) +
      #   geom_text_repel(data = filter(prep_ggplot, variable == max(hours)), aes(label = gene_celltype, color = gene_celltype), direction = "y", hjust = -0.2, cex = 3.5, segment.color = "grey50", xlim = c(-100, 100))+ 
      #   labs(title = paste0(unique(celltype_combinations$ligand), " -> ", unique(celltype_combinations$target), " (weight: ", unique(celltype_combinations$weight), ")"), x = "hours", y = "logCPM") 
      #   
      # 
      # ## 2) Heatmap type
      } else{
      print("--------Skipped NicheNet analysis: length(Gene set of interest) < 10")
    }
  }
}

saveRDS(ligand_target_list, file = "C:/Users/pc/Desktop/MastersProject/Dynamic_NicheNet/Dynamic_NicheNet/Dynamic_NicheNet_PerTime/ligand_target_list.rds")
      

```